---
title: "Assignment 2"
author: "Dmitrii Bychkov / 014059377" 
output: html_document
---

#  Cross sectional study

Tehty: `r format(Sys.time()) `

### Instruction

Complete this assignment file and return it and resulting html-file to moodle.

Write as "author:" your name and student number (if you have one)


In this assignment we investigate association between HDL (`hdl`) (y-variable) and the following variables:


 * Assignment (`age`)
 * height (`height.cm`)
 * weight (`weight.kg`)
 * sex (`gender`)
 * loaction (`location`)
 * ratio.ter


## Read data

We use [diabetes data](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/diabetes.html)


```{r}

# Load example diabetes data, this is originally SPSS file
# Use package "foreign" 
library(foreign)
load(file=url("http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/diabetes.sav"))

# First 2 rows
head(diabetes,2)

# Simple summaries
summary(diabetes)

# New variable waist/hip ratio in tertiles
with(diabetes,table(cut(ratio,quantile(ratio,probs=c(0,1/3,2/3,1),na.rm=TRUE),
                        include.lowest=TRUE),useNA="ifany"))

diabetes$ratio.ter<-with(diabetes,cut(ratio,quantile(ratio,probs=c(0,1/3,2/3,1),na.rm=TRUE),
                        include.lowest=TRUE))

```


## Variable transformations

Make new variables:

 * `height.cm` height in inches transformed to cm
 * `weight.kg` pounds into kg


```{r}

diabetes$height.cm  <- diabetes$height * 2.54
diabetes$weight.kg  <- diabetes$weight / 2.2

```


## Tabulate

Make tables and output them into html.


Take into account if there is any missing values.


```{r message=FALSE}
library(tables)
```

```{r}
tmp.tab0  <- tabular( (gender) ~ (1+Heading(Complete)*complete.cases(diabetes))*(hdl)*((n=1)+mean+sd+median),
        data=diabetes )


print(tmp.tab0)
```

__Since diabetes type II is assosiated with obesity, let's take a look at _weight_ and _waist_; But first we need to pre-process columns:__

```{r}
diabetes$weight.Q <- with(diabetes, cut(weight.kg, 
                                breaks=quantile(weight.kg, probs=seq(0,1, by=0.25), na.rm=TRUE), 
                                include.lowest=TRUE))
diabetes$waist.Q <- with(diabetes, cut(waist, 
                                breaks=quantile(waist, probs=seq(0,1, by=0.25), na.rm=TRUE), 
                                include.lowest=TRUE))

```

__weight:__
```{r}
tmp.tab1  <- tabular( (weight.Q) ~ (1+Heading(Complete)*complete.cases(diabetes))*(ratio)*((n=1)+mean+sd+median),
        data=diabetes)
print(tmp.tab1)
```

__waist:__
```{r}
tmp.tab2  <- tabular( (waist.Q) ~ (1+Heading(Complete)*complete.cases(diabetes))*(ratio)*((n=1)+mean+sd+median),
        data=diabetes)
print(tmp.tab2)
```

Output html tables.

```{r, results='asis',echo=FALSE}
html(tmp.tab1)
html(tmp.tab2)
```

__So we can see as the weight (and the waist of course) of patient grows the cholesterol/HDL ration increases accordingly. Seems logical.__

## Make graphs


Make scatterplots, boxplots, and other types of graphs for description.
```{r, message=FALSE}
library(ggplot2)
library(reshape2)
library(broom)
```
__We need to melt the data first:__
```{r, error=FALSE, message=FALSE, warning=FALSE}
melted <- melt(diabetes[(diabetes$ratio < 10 & diabetes$bp.1s < 200), c('id','bp.1s','bp.1d','weight.kg')], 
               measure.vars = c('bp.1s','bp.1d'), na.rm = TRUE)
```

__Is there some dependency between blood pressure and patient's weight? __
```{r, warning=FALSE, message=FALSE}
ggplot(melted, aes(x=weight.kg, y=value, colour=variable)) +
  geom_point() + geom_smooth(method = 'lm')
```

__There is some, but hard to say from the plot how significant it is__

```{r, warning=FALSE, message=FALSE}
ggplot(na.omit(diabetes), aes(x=location, y=ratio, colour=frame)) +
  geom_boxplot()
```
```{r, warning=FALSE, message=FALSE}
ggplot(na.omit(diabetes), aes(x=gender, y=hdl)) +
  geom_boxplot()
```

## Testing

Test if sex (gender) or location are associated with HDL. 

**NOTE!** These are dichotomic variables


## Linear model

Make linear regression models with HDL as response (y) variable. Select the best model and interpret
results.

Document modeling. Output parameter estimates, p values and confidence intervals of the best model.


```{r}
# Explaining glyhb with weight, make model named tmp.m0
# NOTE! becasue of missing values "subset=complete.cases(diabetes)""
tmp.m0  <- glm( hdl ~ gender, data=diabetes,subset=complete.cases(diabetes) )
#print(tmp.m0)
summary(tmp.m0)
```

__1st Model - Gender & Age__
```{r}
tmp.m1  <- lm( hdl ~ gender + age, data=diabetes,subset=complete.cases(diabetes) )
summary(tmp.m1)
```
__Comments: Both gender an age are assosiated with HDL levels. Significance is not very strong with p-values just a bit below the 5% threshold. Women expose higher levels of HDL as indicated by a coefficient of 7.9. Age is meant to increase HDL levels by about 22% each year. The two factors together explain only about 7.5% of variance observed in the data, which is quite low.__

__2nd Model - Add weight to the 1st model __
```{r}
tmp.m2  <- lm( hdl ~ gender + age + weight + stab.glu, data=diabetes,subset=complete.cases(diabetes) )
summary(tmp.m2)
```
__Comments: Weight appear to be highly significant predictor of HDL levels with effect size of -0.12. Negative means that higher weight is assosiated with lower levels of 'good' cholesterol. Stabilized glucose (stab.glu) is also negatively associated with HDL levels. Second model explains 20% of variance.__

__3rd Model: Introduce interaction term__
```{r}
tmp.m3  <- lm( hdl ~ gender + age*weight + stab.glu, data=diabetes,subset=complete.cases(diabetes) )
summary(tmp.m3)
```
__Comments: By assuming interaction between weight and age we improved R-squared up to almost 28 %.__

__Now let's compare the last to models (with and without an interaction term)__
```{r}
anova(tmp.m2, tmp.m3, test='Chisq')
```
__Anova suggests significant difference in the residuals of the two models. So we should prefer tmp.m3 model as it explainc more variance with the same set of parameters. __

__Resulting model coefficients:__
```{r}
tmp.m3$coefficients
```

__P-values are found from summary() output under Pr(>|t|) for each parameter:__
```{r}
summary(tmp.m3)
```
__95% Confiddence intervals for each of the coefficients:__
```{r}
confint(tmp.m3)
```